---
title: "Communicating Data"
output: html_notebook
---

```{r}
library(lubridate)
library(tsibble)
library(tsibbledata)
library(tidyverse)
library(slider)
library(geosphere)


nyc_bikes_df <- nyc_bikes
```

1 Business questions
In your report, please consider the following questions:

What is the pattern of bike hires over time (e.g. within a year, month, week, or day)?

Do bike hire patterns differ between bike rider demographics? (e.g. gender, type of trip, age)

Any other insights?

For every visualisation you create, consider whether and how it can provide actionable insights to the business (in this case NYC Citi Bike, a bike share operator) and how it relates to key performance indicators (in this case the overall objective is to increase bike hires). Also, remember to include a description and interpretation of each plot so as to explain the context and insights.

```{r}
glimpse(nyc_bikes_df) 


```
```{r}
index_var(nyc_bikes_df)
```

```{r}


nyc_bikes_df <- nyc_bikes_df %>% 
  mutate(date = date(start_time), week = week(start_time), month = month(start_time, label = TRUE, abbr = FALSE), year = year(start_time)) %>% 
  mutate(day_length = daylength(start_lat, date)) %>% 
  mutate(
    distance =  distGeo(
                    matrix(c(start_long, start_lat), ncol = 2),                     matrix(c(end_long, end_lat), ncol = 2)
                        )
        )
```

```{r}
#bike hire pattern within year
nyc_bikes_df %>% 
index_by(week) %>% 
  summarise(
    mean_distance = mean(distance)
  ) %>% 
ggplot() +
  aes(x = week, y = mean_distance)+
  geom_line()

```
```{r}
nyc_bikes_df %>% 
index_by(day) %>% 
  summarise(
   no_of_hires = n()
  ) %>% 
ggplot() +
  #aes(x = week)+
  #geom_line(aes(y = no_of_hires))+
  geom_line(aes(x = no_of_hires, y = day_length, colour = "red"))
```

```{r}
bike_distance <- nyc_bikes_df %>% 
  select(bike_id, distance) %>% 
group_by(bike_id) %>% 
  summarise(
    mean_distance = sum(distance)
  )
```

