---
title: "Queries Part 1 - Filtering with `WHERE`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
    css: ../../../styles.css
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center')
```

# Learning Objectives  

* Be able to use additional comparison operators, and `AND` and `OR` combinations
* Write queries using `BETWEEN`, `NOT` and `IN`
* Understand and be able to use inexact comparisons with `LIKE` and wildcards
* Understand `IS NULL`

**Duration - 90 minutes**  

# Introduction

As stated earlier, data analysts more often read records in databases than create, update or delete them (recall CRUD). So, it makes sense to focus effort on learning `SELECT` syntax beyond the simple `WHERE` clauses we saw in the last lesson.  
  
We'll work through a series of examples using a database storing employee information for the world-spanning 'OmniCorp' corporation.   

As you will see, the most difficult task faced by data analyst is often to take a 'business question', framed in the context and logic of an organisation, and render it into a 'data question', i.e. a question that can be answered by interrogating data. It's a deeply practical skill that improves with experience.  

Let's get started!  

# Setup

First, we need to create a connection to the database in `DBeaver`:

<br>
<blockquote class="task">
**Task - 10 mins**

Create a connection in `DBeaver` using the following details:

| Connection setting | Value |
|---|---|
| Username | omni_user |
| Password | in2z00yfv3as3hgo |
| Host | db-postgresql-lon1-94437-do-user-4831272-0.a.db.ondigitalocean.com |
| Port | 25061 |
| Database | omni_pool |

```{r, eval=TRUE, include=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
library(RPostgres)

db = dbConnect(
  Postgres(), 
  user = 'omni_user',
  password = 'in2z00yfv3as3hgo',
  dbname = 'omni_pool',
  host = 'db-postgresql-lon1-94437-do-user-4831272-0.a.db.ondigitalocean.com',
  port = 25061,
  sslmode = 'require'
)
```

Check that your connection is working by exploring the `employees` table, and then writing and executing a query to return all records from the same table.

<details>
<summary>**Solution**</summary>

```{sql, connection = db, output.var="out"}
SELECT * 
FROM employees;
```
```{r, echo=FALSE, eval=TRUE}
out
```
</details>
</blockquote>
<br>

# Simple `WHERE` clauses  

So far we've seen pretty simple `WHERE` clauses, e.g. 

<br>
<center>
**"Find the employee with `id` equal to 3."**
</center>
<br>

```{sql, connection = db, output.var="out"}
SELECT * 
FROM employees 
WHERE id = 3;
```
```{r, echo=FALSE, eval=TRUE}
out
```

# Additional comparison operators  

OK, but what about this problem:  

<br>
<center>
**"Find all the employees working 0.5 full-time equivalent hours or greater."**
</center>
<br>

We can solve this using operators other than `=`    

| Operator | Meaning |
| --- | --- |
| `!=` | not equal to |
| `>` | greater than |
| `<` | less than |
| `>=` | greater than or equal to |
| `<=` | less than or equal to |

The correct operator from above is `>=` (greater than or equal to):  

```{sql, connection = db, output.var="out"}
SELECT * 
FROM employees 
WHERE fte_hours >= 0.5;
```
```{r, echo=FALSE, eval=TRUE}
out
```

<br>
<blockquote class="task">
**Task - 2 mins** 

Write and execute a query answering this problem:  

<br>
<center>
**"Find all the employees *not* based in Brazil."**
</center>
<br>

<details>
<summary>**Solution**</summary>

```{sql, connection = db, output.var="out"}
SELECT * 
FROM employees 
WHERE country != 'Brazil';
```
```{r, echo=FALSE, eval=TRUE}
out
```

</details>
</blockquote>
<br>


# `AND` and `OR` 

If required, we can create more complex clauses using the `AND` and `OR` operators

<br>
<center>
**"Find all employees in China who started working for OmniCorp in 2019."**
</center>
<br>

```{sql, connection = db, output.var="out"}
SELECT * 
FROM employees 
WHERE country = 'China' AND start_date >= '2019-01-01' AND start_date <= '2019-12-31';
```
```{r, echo=FALSE, eval=TRUE}
out
```

Sometimes we have to be careful with the **order of evaluation** of conditions. Consider the following example:

<br>
<center>
**"Of all the employees based in China, find those who either started working for OmniCorp from 2019 onwards or who are enrolled in the pension scheme."**
</center>
<br>

The logic of this as written is fairly clear. All the returned employees should be based in China **and** they should either have started work for OmniCorp from 2019 onwards **or** be enrolled in the pension scheme. Let's have a first go at writing this!

```{sql, connection = db, output.var="out"}
SELECT * 
FROM employees 
WHERE country = 'China' AND start_date >= '2019-01-01' OR pension_enrol = TRUE;
```
```{r, echo=FALSE, eval=TRUE}
out
```

Argh! We get employees from countries other than China! What's gone wrong? We need to worry about the **order of evaluation**. We want the `OR` operation to execute **before** the `AND` operation, and we can enforce this by use of parentheses!

```{sql, connection = db, output.var="out"}
SELECT * 
FROM employees 
WHERE country = 'China' AND (start_date >= '2019-01-01' OR pension_enrol = TRUE);
```
```{r, echo=FALSE, eval=TRUE}
out
```

That's better! Let's see a few more examples using `AND` and `OR`

<br>
<center>
**"Find all employees who work between 0.25 and 0.5 full-time equivalent hours inclusive."**
</center>
<br>

```{sql, connection = db, output.var="out"}
SELECT * 
FROM employees 
WHERE fte_hours >= 0.25 AND fte_hours <= 0.5;
```
```{r, echo=FALSE, eval=TRUE}
out
```

<br>
<center>
**"Find all employees who started working for OmniCorp in years other than 2017."**
</center>
<br>

```{sql, connection = db, output.var="out"}
SELECT * 
FROM employees 
WHERE start_date < '2017-01-01' OR start_date > '2017-12-31';
```
```{r, echo=FALSE, eval=TRUE}
out
```

# `BETWEEN`, `NOT` and `IN`

The syntax in the last two examples is clumsy: it would be better to be able to define a range and then `SELECT` for records with fields in/not in that range. The `BETWEEN` keyword lets us do just that!  

Let's rewrite the previous two queries using `BETWEEN`:  

```{sql, connection = db, output.var="out"}
SELECT * 
FROM employees 
WHERE fte_hours BETWEEN 0.25 AND 0.5;
```
```{r, echo=FALSE, eval=TRUE}
out
```

```{sql, connection = db, output.var="out"}
SELECT * 
FROM employees 
WHERE start_date NOT BETWEEN '2017-01-01' AND '2017-12-31';
```
```{r, echo=FALSE, eval=TRUE}
out
```

Note these two points:

* the range defined by `BETWEEN` is **inclusive** of the end points. So, in the first example, records with `fte_hours` of exactly 0.25 or 0.5 will be selected
* we use the `NOT BETWEEN` combination in the second example to select all records with `start_date` not in the range (where, again, the end points are included in the range).

<br>
<blockquote class="task">
**Task - 5 mins** 

See if you can work out the query syntax to answer this problem:

<br>
<center>
**"Find all employees who started work at OmniCorp in 2016 who work 0.5 full time equivalent hours or greater."**
</center>
<br>

<details>
<summary>**Hint**</summary>  
You need an `AND` combination of two conditions, one of which involves `BETWEEN`
</details>

<details>
<summary>**Solution**</summary>

```{sql, connection = db, output.var="out"}
--parentheses might help you to see different logical 'levels' of the conditions
SELECT * 
FROM employees 
WHERE (start_date BETWEEN '2016-01-01' AND '2016-12-31') AND (fte_hours >= 0.5);
```
```{r, echo=FALSE, eval=TRUE}
out
```
</details>
</blockquote>
<br>

The `IN` operator helps us deal with discrete valued fields more efficiently. Consider:

<br>
<center>**"Find all employees based in Spain, South Africa, Ireland or Germany."**
</center>
<br>

The long-winded way to write this is:  

```{sql, connection = db, output.var="out"}
SELECT * 
FROM employees 
WHERE country = 'Spain' OR country = 'South Africa' OR country = 'Ireland' OR country = 'Germany';
```
```{r, echo=FALSE, eval=TRUE}
out
```

or using `IN`:  

```{sql, connection = db, output.var="out"}
SELECT * 
FROM employees 
WHERE country IN ('Spain', 'South Africa', 'Ireland', 'Germany');
```
```{r, echo=FALSE, eval=TRUE}
out
```

We can also use `NOT` with `IN`:  

<br>
<center>**"Find all employees based in countries other than Finland, Argentina or Canada."**
</center>
<br>

```{sql, connection = db, output.var="out"}
SELECT * 
FROM employees 
WHERE country NOT IN ('Finland', 'Argentina', 'Canada');
```
```{r, echo=FALSE, eval=TRUE}
out
```

# `LIKE` and wildcards

Your manager comes to you and says:

<br>
<center>**"I was talking with a colleague from Greece last month, I can't remember their last name exactly, I think it began 'Mc...' something-or-other. Can you find them?"**
</center>
<br>

You can see we're dealing with an **inexact comparison** here. How do we do this in `SQL`? Using the `LIKE` operator with a **wildcard**.  

```{sql, connection = db, output.var="out"}
SELECT * 
FROM employees 
WHERE country = 'Greece' AND last_name LIKE 'Mc%';
```
```{r, echo=FALSE, eval=TRUE}
out
```

The string after `LIKE` is called a **pattern**, so in this case, the pattern is `'Mc%``

Here are the wildcards we can use in a pattern:  

| Wildcard | Meaning |
| --- | --- |
| _ | a single character |
| % | zero or more characters |

We can place wildcards **anywhere** inside the pattern:  

<br>
<center>**"Find all employees with last names containing the phrase 'ere' anywhere."**
</center>
<br>

```{sql, connection = db, output.var="out"}
SELECT * 
FROM employees 
WHERE last_name LIKE '%ere%';
```
```{r, echo=FALSE, eval=TRUE}
out
```

<br>
<center>**"Find all employees in the Legal department with a last name beginning with 'D'."**
</center>
<br>

```{sql, connection = db, output.var="out"}
SELECT * 
FROM employees 
WHERE department = 'Legal' AND last_name LIKE 'D%';
```
```{r, echo=FALSE, eval=TRUE}
out
```

<br>
<blockquote class="task">
**Task - 2 mins** 

Write a query using `LIKE` and wildcards to answer:
  
<br>
<center>**"Find all employees having 'a' as the second letter of their first names."**
</center>
<br>
  
<details>
<summary>**Hint**</summary> 
You can use a '_' wildcard for the first letter of `first_name`.  
</details>
  
<details>
<summary>**Solution**</summary>

```{sql, connection = db, output.var="out"}
SELECT * 
FROM employees 
WHERE first_name LIKE '_a%';
```
```{r, echo=FALSE, eval=TRUE}
out
```

</details>
</blockquote>
<br>

`LIKE` distinguishes between capital and lower case letters. If we need a case-insensitive version, we can use `ILIKE`:

<br>
<center>**"Find all employees whose last name contains the letters 'ha' anywhere."**
</center>
<br>

```{sql, connection = db, output.var="out"}
SELECT * 
FROM employees 
WHERE last_name ILIKE '%ha%';
```
```{r, echo=FALSE, eval=TRUE}
out
```

# `IS NULL`

<br>
<center>**"We need to ensure our employee records are up-to-date. Find all the employees who do not have a listed email address."**
</center>
<br>

We can use the `IS NULL` condition to fulfill this request:

```{sql, connection = db, output.var="out"}
SELECT * 
FROM employees 
WHERE email IS NULL;
```
```{r, echo=FALSE, eval=TRUE}
out
```

We can use the `NOT` operator here too: `IS NOT NULL` is a valid condition!

# Recap - Full query syntax

In summary, here are the different keyword components of a `SELECT` query, the order in which they must appear, and whether they are required or optional <br>

This table shows where we got to in this lesson:  

| Order | Keyword | Specifies | Required? |
| --- | --- |--- | --- |
| 1 | SELECT | Column to query | Yes |
| 2 | FROM | Table to query |  Yes |
| 3 | WHERE | Row-level filter | No |


and here's a look at what's still to come:  

| Order | Keyword | Specifies | Required? |
| --- | --- |--- | --- |
| 5 | GROUP BY| Grouping for aggregates | No |
| 6 | HAVING | Group-level filter | No |
| 7 | ORDER BY | Sort order | No |
| 8 | LIMIT | How many records to return | No |






