---
title: "SQL Day 1 Lab and Homework"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
    css: ../../../styles.css
  pdf_document: default
---


```{r, eval=TRUE, include=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(RPostgres)

db = dbConnect(
  Postgres(), 
  user = 'omni_user',
  password = 'in2z00yfv3as3hgo',
  dbname = 'omni_pool',
  host = 'db-postgresql-lon1-94437-do-user-4831272-0.a.db.ondigitalocean.com',
  port = 25061,
  sslmode = 'require',
  bigint = "numeric"
)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center')
```
  
**Duration - 120 minutes**  
  
In these practice questions you'll return again to the `omni_pool` database, so re-establish your connection in `DBeaver` if it has lapsed. 

<br>
<div class='emphasis'>
**Submission**

When answering these questions, please save your `SQL` script in `DBeaver`, copy it over to your homework repo and then add, commit and push that for submission tonight. Please add section and question numbers as `SQL` comments, like so:

```{sql, eval=FALSE}
/* MVP */
/* Q1 */
```
</div>
<br>

# MVP Questions

1. Find all the employees who work in the 'Human Resources' `department`.
  
2. Get the `first_name`, `last_name`, and `country` of the employees who work in the 'Legal' `department`.
  
3. Count the number of employees based in Portugal.
  
4. Count the number of employees based in either Portugal or Spain.
  
5. Count the number of pay_details records lacking a `local_account_no`.
  
<details>
<summary>**Hint**</summary>
Run a query first to get some records in the `pay_details` table, just to see what the data looks like
</details>
  
6. Get a table with employees `first_name` and `last_name` ordered alphabetically by `last_name` (put any `NULL`s last).
  
7. How many employees have a `first_name` beginning with 'F'?
  
8. Count the number of pension enrolled employees not based in either France or Germany.
  
9. Obtain a count by `department` of the employees who started work with the corporation in 2003.  
  
10. Obtain a table showing `department`, `fte_hours` and the number of employees in each department who work each `fte_hours` pattern. Order the table alphabetically by `department`, and then in ascending order of `fte_hours`. 

<details>
<summary>**Hint**</summary>
You need to `GROUP BY` two columns here.
</details>

11. Obtain a table showing any `department`s in which there are two or more employees lacking a stored first name. Order the table in descending order of the number of employees lacking a first name, and then in alphabetical order by `department`.

12. [**Tough!**]  Find the proportion of employees in each department who are `grade` 1.

<details>
<summary>**Hints**</summary>

* Think of the desired proportion for a given `department` as the number of employees in that department who are `grade` 1, divided by the total number of employees in that department.  
* You can write an expression in a `SELECT` statement, e.g. `grade = 1`. This would result in `BOOLEAN` values.  
* If you could convert `BOOLEAN` to `INTEGER` 1 and 0, you could sum them. The `CAST()` function lets you convert data types.  
* In `SQL`, an `INTEGER` divided by an `INTEGER` yields an `INTEGER`. To get a `REAL` value, you need to convert the top, bottom or both sides of the division to `REAL`.  

</details>
  
# Extension Questions

Some of these problems may need you to do some online research on `SQL` statements we haven't seen in the lessons today... Don't worry, we'll give you pointers, and it's good practice looking up help and answers online; data analysts and programmers do this all the time! If you get stuck, it might also help to sketch out a rough version of the table you want on paper (the column headings and first few rows are usually enough). 

1. Do a count by **year** of the `start_date` of all employees, ordered most recent year last.  
  
<details>
<summary>**Hint**</summary>
Investigate how to use the `SQL` `EXTRACT()` function to get the year from a date.
</details>
  
2. Return the `first_name`, `last_name` and `salary` of all employees together with a new column called `salary_class` with a value `'low'` where `salary` is less than 40,000 and value `'high'` where `salary` is greater than or equal to 40,000.  

<details>
<summary>**Hints**</summary>

*  Investigate how to use a `SQL` `CASE` statement to return the required values `'low'` and `'high'` based on the value of `salary`.  
*  Think carefully how to deal with `NULL` salaries.  

</details>
  
3. [**Tough!**] The first two digits of the `local_sort_code` (e.g. digits `97` in code `97-09-24`) in the `pay_details` table are indicative of the region of an account. Obtain counts of the number of `pay_details` records bearing each set of first two digits? Make sure that the count of `NULL` `local_sort_codes` comes at the top of the table, and then order all subsequent rows first by counts in descending order, and then by the first two digits in ascending order  Your final table should look something like:

| first_two_digits | count_records |
|---|---|
| `NULL` | 15 |
| 45 | 21 |
| 12 | 20 |
| 24 | 20 |
| 56 | 20 |
| 08 | 18 |
| 34 | 18 |
| 87 | 17 |
| ... | ... |

<details>
<summary>**Hints**</summary>

* This is a really tough problem for day 1 of `SQL`!  
* Investigate the `SUBSTRING()` function to extract the first two digits.  
* Ordering to put `NULL`s in `first_two digits` at the top of the table, but then immediately reverting to ordering by `count_records` may require you to write an expression using a `CASE` statement in `ORDER BY` (you're trying to create a first level of ordering that puts `NULL`s first, but then doesn't care about the ordering of subsequent rows).  
* The next level of ordering will involve `count_records`, and then finally `first_two_digits`.  

</details>
  
4. [**Tough!**] Return **only the numeric** part of the `local_tax_code` in the `pay_details` table, preserving `NULL`s where they exist in this column.

<details>
<summary>**Hints**</summary>

* This is tricky and fussy: this is true of most string formatting problems!  
* Investigate the `REGEXP_REPLACE` function (search pattern `'\D'` will let you find and replace characters that are **not** a digit, flag `'g'` will let you do so 'globally', i.e. at each position in a string). What do you want to replace non-digit characters with?  

</details>

