---
title: "week6_day_@HOmework_GA"
output: html_notebook
---
```{r}
library(janitor)
library(tidyverse)
```
# MVP
Q1
```{r}
phone_data <- read_csv("data/20190928-items.csv") %>% 
  clean_names()
```

```{r}
head(phone_data)
glimpse(phone_data)
```
Q2
```{r}
phone_data %>% 
count(brand)
```
Samsung

Q3
```{r}
samsung <- phone_data %>% 
  filter( brand == "Samsung") %>% 
  select(title, rating, total_reviews)
```

```{r}
samsung_stats <- samsung  %>% 
summarise(
    num = n(),
    mean = mean(rating),
    sd = sd(rating)
  )
samsung_stats
```

```{r}
samsung %>%
  ggplot(aes(x = rating)) +
  geom_histogram(aes(y = ..density..), col = "white", bins = 25)
```

```{r}
samsung %>%
  ggplot(aes(x = rating)) +
  geom_histogram(aes(y = ..density..), col = "white", bins = 25) +
  stat_function(
    fun = dnorm, 
    args = list(
      mean = samsung_stats$mean, 
      sd = samsung_stats$sd
    ),
    col = "red"
  )  
```
Normal distribution does seem to be a fairly good model although there is perhaps evidence of bimodality around 5 as a rating 


Q4
```{r}
samsung %>%
  filter(rating >= samsung_stats$mean - samsung_stats$sd) %>%
  filter(rating <= samsung_stats$mean + samsung_stats$sd) %>%
  summarise(prop_within_1sd = n() / nrow(samsung))

samsung %>%
  filter(rating >= samsung_stats$mean - 2 * samsung_stats$sd) %>%
  filter(rating <= samsung_stats$mean + 2 * samsung_stats$sd) %>%
  summarise(prop_within_2sd = n() / nrow(samsung))

samsung %>%
  filter(rating >= samsung_stats$mean - 3 * samsung_stats$sd) %>%
  filter(rating <= samsung_stats$mean + 3 * samsung_stats$sd) %>%
  summarise(prop_within_3sd = n() / nrow(samsung))
```



Ext
What is a ‘quantile-quantile plot’?
Teststo check if 2 data sets fit the same distribution

What do we expect to see in a qqnorm() plot if data is nearly normally distributed?
plot is roughly a 45degree staright line from the origin
```{r}

```

