---
title: "Queries Part 2 - Manipulating Returned Data and Grouping"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
    css: ../../../styles.css
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center')
```

# Learning Objectives<br>

* Be able to create column aliases using `AS`
* Use `DISTINCT` to return unique records by column
* Understand and be able to use aggregate functions
* Be able to sort records and limit the number returned
* Understand `GROUP BY` and `HAVING` for group level aggregation and filtering

**Duration - 2 hours 30 minutes**<br>

This is a continuation of the last in which we learned how to filter data at the row level using the `WHERE` clause. Now we'll add to our query toolkit with some more advanced syntax! 

# Setup (again)...

Re-establish the connection to the `omni_pool` database in `DBeaver` (you may not have to do this if you ran a query recently).

```{r, eval=TRUE, include=FALSE, message=FALSE, warning=FALSE}
library(RPostgres)

db = dbConnect(
  Postgres(), 
  user = 'omni_user',
  password = 'in2z00yfv3as3hgo',
  dbname = 'omni_pool',
  host = 'db-postgresql-lon1-94437-do-user-4831272-0.a.db.ondigitalocean.com',
  port = 25061,
  sslmode = 'require',
  bigint = "numeric"
)
```

# Manipulating returned data

After we have applied conditions to filter data in a `SELECT` statement, we can also manipulate what is returned. The easiest way to do this is by limiting the returned columns (fields)!

```{sql, connection = db, output.var="out"}
SELECT 
  id, 
  first_name, 
  last_name 
FROM employees 
WHERE department = 'Accounting';
```
```{r, echo=FALSE, eval=TRUE}
out
```


So here we return only the `id`, `first_name` and `last_name` columns.

`SQL` offers additional operators to manipulate the returned data, which we will examine in this lesson.

# Aliases via `AS`

<br>
<center>
**"Can we get a list of all employees with their first and last names combined together into one field called 'full_name'?"**
</center>
<br>

**Column aliases** are the way to solve problems like these! First, we use the `CONCAT()` function to **concatenate** (this is just a fancy way of saying 'join strings together') each pair of names into the full name. Next, we set up a column alias using `AS full_name` to store the concatenated strings. 

Let's see this in operation first with some strings where the contents are known:

```{sql, connection = db, output.var="out"}
SELECT
  CONCAT('Hello', ' ', 'there!') AS greeting;
```
```{r, echo=FALSE, eval=TRUE}
out
```

You can also use the `||` operator for this:

```{sql, connection = db, output.var="out"}
SELECT
  'Hello' || ' ' || 'there!' AS greeting;
```
```{r, echo=FALSE, eval=TRUE}
out
```

We concatenated `'Hello'` with a space `' '` and `'there!'`, and named the column `greeting`. You can always try things like this in `SQL` to explore and help you understand syntax and functions. Now we try a similar approach with the columns in our table:

```{sql, connection = db, output.var="out"}
SELECT 
  id, 
  first_name, 
  last_name, 
  CONCAT(first_name, ' ', last_name) AS full_name 
FROM employees;
```
```{r, echo=FALSE, eval=TRUE}
out
```

The new `full_name` column will be at the right of the output. We see a problem with this: some of the records now have single names for `full_name`, but this represents a problem with the underlying data. We could add in a `WHERE` clause to filter out these problem rows.


<br>
<blockquote class="task">
**Task - 2 mins** 

Add a `WHERE` clause to the query above to filter out any rows that don't have both a first and second name.  

<details>
<summary>**Hint**</summary>
You need an `AND` combination of two `IS NOT NULL` conditions
</details>

<details>
<summary>**Solution**</summary>
```{sql, connection = db, output.var="out"}
SELECT 
  id, 
  first_name,
  last_name,
  CONCAT(first_name, ' ', last_name) AS full_name 
FROM employees 
WHERE first_name IS NOT NULL AND last_name IS NOT NULL;
```
```{r, echo=FALSE, eval=TRUE}
out
```
</details>
</blockquote>
<br>

It is good practice to use aliases when creating new columns or using aggregate functions (coming up soon), so that if someone else (including your future self) examines your `SQL` output, the columns will have useful names. To see what happens without an alias, let's run the query again without `AS full_name`:

```{sql, connection = db, output.var="out"}
SELECT 
  id, 
  first_name,
  last_name,
  CONCAT(first_name, ' ', last_name)
FROM employees 
WHERE first_name IS NOT NULL AND last_name IS NOT NULL;
```
```{r, echo=FALSE, eval=TRUE}
out
```

# `DISTINCT()`

<br>
<center>
**"Our database may be out of date! After the recent restructuring, we should now have six departments in the corporation. How many departments do employees belong to at present in the database?"**
</center>
<br>

How do we solve this problem? It's not enough to simply return **all** the departments of the employees, as there will be a large amount of duplication. Instead, we need the **unique** list of departments. The `DISTINCT()` function returns a unique list.

```{sql, connection = db, output.var="out"}
SELECT 
  DISTINCT(department) 
FROM employees;
```
```{r, echo=FALSE, eval=TRUE}
out
```


# Aggregate functions

<br>
<center>
**"How many employees started work for the corporation in 2001?"**
</center>
<br>

The `COUNT()` aggregate function can help us with counting problems:

```{sql, connection = db, output.var="out"}
SELECT 
  COUNT(*) 
FROM employees 
WHERE start_date BETWEEN '2001-01-01' AND '2001-12-31';
```
```{r, echo=FALSE, eval=TRUE}
out
```

In addition to `COUNT()`, we have the following aggregate functions:
<br>

| Function | Purpose |
| --- | --- |
| `SUM()` | sum of a column |
| `AVG()` | average of a column |
| `MIN()` | minimum value of a column |
| `MAX()` | maximum value of a column |

As mentioned above it's good practice to use an alias (using the `AS` function) to give meaning to the result, such as:

```{sql, connection = db, output.var="out"}
SELECT 
  COUNT(*) AS total_employ_2001
FROM employees 
WHERE start_date BETWEEN '2001-01-01' AND '2001-12-31';
```
```{r, echo=FALSE, eval=TRUE}
out
```


<br>
<blockquote class="task">
**Task - 10 mins** 

Design queries using aggregate functions and what you have learned so far to answer the following questions:<br>

<br>
<center>
**1. "What are the maximum and minimum salaries of all employees?"**

**2. "What is the average salary of employees in the Human Resources department?"**

**3. "How much does the corporation spend on the salaries of employees hired in 2018?"**
</center>
<br>

<details>
<summary>**Solution**</summary>
1.  

```{sql, connection = db, output.var="out"}
SELECT 
  MAX(salary) AS max_salary
FROM employees;
```
```{r, echo=FALSE, eval=TRUE}
out
```

```{sql, connection = db, output.var="out"}
SELECT 
  MIN(salary) AS min_salary
FROM employees;
```
```{r, echo=FALSE, eval=TRUE}
out
```

We can also do this in a single query:  

```{sql, connection = db, output.var="out"}
SELECT 
  MAX(salary) AS max_salary,
  MIN(salary) AS min_salary
FROM employees;
```
```{r, echo=FALSE, eval=TRUE}
out
```

2.  

```{sql, connection = db, output.var="out"}
SELECT 
  AVG(salary)as avg_sal
FROM employees 
WHERE department = 'Human Resources';
```
```{r, echo=FALSE, eval=TRUE}
out
```

3.
```{sql, connection = db, output.var="out"}
SELECT 
  SUM(salary) AS total_sals_2018
FROM employees 
WHERE start_date BETWEEN '2018-01-01' AND '2018-12-31';
```
```{r, echo=FALSE, eval=TRUE}
out
```
</details>
</blockquote>
<br>

# Sorting by columns

The `ORDER BY` operator lets us **sort** the returns of queries, either in descending (`DESC`) or ascending (`ASC`) order. The `ORDER BY` operator and associated keywords **always comes after** any `WHERE` clause.

The `LIMIT` operator is a natural partner to `ORDER BY`: it lets us limit **how many** records are returned by a query.

So, we saw before the minimum and maximum salaries of employees in the corporation. Let's find out which employees earn those salaries using the new operators!

```{sql, connection = db, output.var="out"}
--First, the minimum salaried employee. 
--Order employees by salary in *ascending* order, return the first, 
--We need to be careful with NULLs here, let's filter out employees with NULL salary
SELECT * 
FROM employees 
WHERE salary IS NOT NULL 
ORDER BY salary ASC 
LIMIT 1;
```
```{r, echo=FALSE, eval=TRUE}
out
```

```{sql, connection = db, output.var="out"}
--Next, the maximum salaried employee. 
--Order employees by salary in *descending* order, return the first 
SELECT * 
FROM employees 
WHERE salary IS NOT NULL 
ORDER BY salary DESC 
LIMIT 1;
```
```{r, echo=FALSE, eval=TRUE}
out
```

You see above we had problems in ordering with `NULL`s in the column we're sorting on. We can either filter them out with a `WHERE` clause, as we did above, or we can use the operators `NULLS FIRST` or `NULLS LAST` to specify where to put them in the list of records. These operators are always placed immediately after `DESC` or `ASC` for the respective column.

Let's rewrite the queries above using these operators:

```{sql, connection = db, output.var="out"}
SELECT * 
FROM employees 
ORDER BY salary ASC NULLS LAST 
LIMIT 1;
```
```{r, echo=FALSE, eval=TRUE}
out
```

```{sql, connection = db, output.var="out"}
SELECT * 
FROM employees 
ORDER BY salary DESC NULLS LAST 
LIMIT 1;
```
```{r, echo=FALSE, eval=TRUE}
out
```

Finally, we can perform **multi-level sorts** on two or more columns:

<br>
<center>
**"Order employees by full-time equivalent hours, highest first, and then alphabetically by last name."**
</center>
<br>

```{sql, connection = db, output.var="out"}
SELECT * 
FROM employees 
ORDER BY 
  fte_hours DESC NULLS LAST, 
  last_name ASC NULLS LAST;
```
```{r, echo=FALSE, eval=TRUE}
out
```

<br>
<blockquote class="task">
**Task - 5 mins** 

Write queries to answer the following questions using the operators introduced in this section.

<br>
<center>
**1. "Get the details of the longest-serving employee of the corporation."**
  
**2. "Get the details of the highest paid employee of the corporation in Libya."**
</center>
<br>

<details>
<summary>**Solution**</summary>

1.  

```{sql, connection = db, output.var="out"}
SELECT * 
FROM employees 
ORDER BY start_date ASC NULLS LAST 
LIMIT 1;
```
```{r, echo=FALSE, eval=TRUE}
out
```

2.  

```{sql, connection = db, output.var="out"}
SELECT * 
FROM employees 
WHERE country = 'Libya' 
ORDER BY salary DESC NULLS LAST 
LIMIT 1;
```
```{r, echo=FALSE, eval=TRUE}
out
```

</details>
</blockquote>
<br>

# `GROUP BY`

Consider the following problem

<br>
<center>
**"Find the number of employees within each department of the corporation."**
</center>
<br>

We could solve this manually, but it would be a real pain. First, we would need to get a list of the departments in the corporation (we saw how to do this earlier using `DISTINCT`), and then write a query using a `COUNT()` aggregate with a `WHERE` clause specifying each department in turn, something like:

```{sql, connection = db, output.var="out"}
SELECT 
  COUNT(id) AS number_employees 
FROM employees 
WHERE department = 'Legal';
```
```{r, echo=FALSE, eval=TRUE}
out
```

and so on 11 more times for each department. Annoying, and not very general `SQL` code: we need to know the 12 departments before we start! Thankfully, `SQL` gives us the `GROUP BY` operator to automate this!

```{sql, connection = db, output.var="out"}
SELECT 
  department, 
  COUNT(id) AS num_employees 
FROM employees 
GROUP BY department;
```
```{r, echo=FALSE, eval=TRUE}
out
```

Yay, this looks more useful! Note what SQL has done here: it first **groups** records by a specified column (`department` in this case), and then applies the aggregate function (`COUNT()`) to **each group**.

Let's see another few examples of grouping in queries:

<br>
<center>
**"How many employees are there in each country?"**
</center>
<br>

```{sql, connection = db, output.var="out"}
SELECT 
  country, 
  COUNT(id) AS num_employees 
FROM employees 
GROUP BY country;
```
```{r, echo=FALSE, eval=TRUE}
out
```

<br>
<center>
**"How many employees in each department work either 0.25 or 0.5 FTE hours?"**
</center>
<br>

```{sql, connection = db, output.var="out"}
--Note the combination of a WHERE clause to filter records and then a GROUP BY. 
--Could also use IN rather than BETWEEN here
SELECT 
  department, 
  COUNT(id) AS num_fte_quarter_half 
FROM employees 
WHERE fte_hours BETWEEN 0.25 AND 0.5 
GROUP BY department;
```
```{r, echo=FALSE, eval=TRUE}
out
```

You'll see here that we are counting the `id` column. You will have seen earlier that we used `COUNT(*)` which counts all rows returned. The difference between the two occurs if there are any `NULL`s in the column you choose for the count. 

Let's take a look at the difference using a column which contains some `NULL`s - the `first_name` column:

```{sql, connection = db, output.var="out"}
SELECT 
  COUNT(first_name) AS count_first_name,
  COUNT(id) AS count_id,
  COUNT(*) AS count_star
FROM employees;
```
```{r, echo=FALSE, eval=TRUE}
out
```


The difference in counts arises because there are `NULL` entries in the `first_name` column, and `COUNT()` skips `NULL`s. Generally speaking, if you want a count of rows, stick to using `COUNT(*)` or `COUNT()` of a primary key (which cannot be `NULL` by definition), e.g. `COUNT(id)` in this case.

<br>
<center>
**"Find the longest time served by any one employee in each department."**
</center>
<br>

This is quite a difficult one! We can use the `NOW()` function to return the current date and time, and then calculate the time served by an employee as `NOW()-start_date`. So, the longest time served by any employee is `NOW()-MIN(start_date)`. This is the aggregate function we will apply to the records grouped by department.

```{sql, connection = db, output.var="out"}
SELECT 
  department, 
  NOW()-MIN(start_date) AS longest_time
FROM employees 
GROUP BY department;
```
```{r, echo=FALSE, eval=TRUE}
out
```

This looks broadly like what we're seeking, but notice that the difference `NOW()-MIN(start_date)` gives a confusingly high amount of precision: days, hours, minutes, seconds etc... Let's just get the days part of this date-time difference. Enter the useful `EXTRACT()` function (let's also see how we could convert this to years):

```{sql, connection = db, output.var="out"}
SELECT 
  department, 
  EXTRACT(DAY FROM NOW()-MIN(start_date)) AS longest_time_days,
  ROUND(EXTRACT(DAY FROM NOW()-MIN(start_date))/365) AS longest_time_approx_years
FROM employees 
GROUP BY department;
```
```{r, echo=FALSE, eval=TRUE}
out
```

Here's the general rule for `GROUP BY`. Once you've grouped rows, your `SELECT` statement is allowed to retrieve the column specified in `GROUP BY`, together with the result of a aggregate function applied across rows in the group. Let's see what happens when we break this rule by trying to `SELECT` the `first_name` in an earlier query:

```{sql, connection = db, eval=FALSE}
SELECT 
  department, 
  first_name,
  COUNT(id) AS num_employees 
FROM employees 
GROUP BY department
```

You'll get an error: `'ERROR: column "employees.first_name" must appear in the GROUP BY clause or be used in an aggregate function'`!

Don't worry if `GROUP BY` takes a while to sink in! It's one of the more difficult concepts in `SQL`, people often struggle a bit with it, but you'll feel more comfortable with practice. Speaking of which...

<br>
<blockquote class="task">
**Task - 10 mins** 

Write two queries using `WHERE` and `GROUP BY` to solve the following problems:

<br>
<center>
**1. "How many employees in each department are enrolled in the pension scheme?"**
  
**2. "Perform a breakdown by country of the number of employees that do not have a stored first name."**
</center>
<br>

<details>
<summary>**Hints**</summary>  
1. we need to apply a `WHERE` clause on `pension_enrol`.<br>
2. remember the `IS NULL` operator.
</details>

<details>
<summary>**Solution**</summary>
1.  

```{sql, connection = db, output.var="out"}
SELECT 
  department, 
  COUNT(id) as num_employees_pension 
FROM employees 
WHERE pension_enrol = TRUE 
GROUP BY department
```
```{r, echo=FALSE, eval=TRUE}
out
```

2.  

```{sql, connection = db, output.var="out"}
SELECT 
  country, 
  COUNT(id) as num_employees_no_first 
FROM employees 
WHERE first_name IS NULL 
GROUP BY country
```
```{r, echo=FALSE, eval=TRUE}
out
```
</details>
</blockquote>
<br>

# `HAVING`

So far we've seen that the `WHERE` clause let's us filter **records**, but what if we wish to filter **groups** by some value of an aggregate function? This is where the `HAVING` operator comes in!

Imagine one of our earlier queries had been even more specific:

<br>
<center>
**"Show those departments in which at least 40 employees work either 0.25 or 0.5 FTE hours"**
</center>
<br>

```{sql, connection = db, output.var="out"}
SELECT 
  department, 
  COUNT(id) as num_fte_quarter_half 
FROM employees 
WHERE fte_hours IN (0.25, 0.5) 
GROUP BY department 
HAVING COUNT(id) >= 40
```
```{r, echo=FALSE, eval=TRUE}
out
```

We've added a `HAVING` clause **after** the `GROUP BY`: notice that it filters using an **aggregate function** applied to a column of the original table.

Let's try one more:

<br>
<center>
**"Show any countries in which the minimum salary amongst pension enrolled employees is less than 21,000 dollars"**
</center>
<br>

```{sql, connection = db, output.var="out"}
SELECT 
  country, 
  MIN(salary) as min_salary 
FROM employees 
WHERE pension_enrol = TRUE 
GROUP BY country 
HAVING MIN(salary) < 21000
```
```{r, echo=FALSE, eval=TRUE}
out
```

Let's break down the two filtering steps:  

* First, **filter the records** to include only pension enrolled employees: `WHERE pension_enrol = TRUE`
* Next, group the filtered records by `country`, and then **filter the groups** to include only countries in which the minimum salary is less than $21,000: `HAVING MIN(salary) < 21000`

<br>
<blockquote class="task">
**Task - 5 mins** 
This is about as tough as `SQL` will get for us in this lesson! Write a query using `WHERE`, `GROUP BY` and `HAVING` to solve the following:

<br>
<center>
**"Show any departments in which the earliest start date amongst grade 1 employees is prior to 1991"**
</center>
<br>

<details>
<summary>**Hints**</summary>  

* You can use `MIN(date)` to get the **earliest** value of a date column
* `WHERE` will involve `grade`
* `GROUP BY` will involve department
* `HAVING` will involve `MIN()` of a column

</details>

<details>
<summary>**Solution**</summary>

```{sql, connection = db, output.var="out"}
SELECT 
  department, 
  MIN(start_date) as earliest_start_date_grade_1_pre_91
FROM employees
WHERE grade = 1
GROUP BY department
HAVING MIN(start_date) < '1991-01-01'
```
```{r, echo=FALSE, eval=TRUE}
out
```
</details>
</blockquote>
<br>

# Subqueries

We're nearly finished, promise! Sometimes we would like to use a value **produced by one query in another query**. This sounds a bit abstract, but it's actually pretty common! As an example, consider this problem:

<br>
<center>
**"Find all the employees in Japan who earn over the company-wide average salary."**
</center>
<br>

We might think a straightforward way to do this is:

```{sql connection = db, eval = FALSE}
SELECT *
FROM employees
WHERE country = 'Japan' AND salary > AVG(salary)
```

but we get `ERROR:  aggregate functions are not allowed in WHERE`! 

In fact, we need to use a **subquery** here. This is just **one query embedded inside another**

```{sql, connection = db, output.var="out"}
SELECT *
FROM employees
WHERE country = 'Japan' AND salary > (SELECT AVG(salary) FROM employees)
```
```{r, echo=FALSE, eval=TRUE}
out
```

If it helps, we can think of 'building this query up' in two steps. First, we write what we want to achieve using a placeholder name 'average_salary':  

```{sql, eval=FALSE}
SELECT *
FROM employees
WHERE country = 'Japan' AND salary > average_salary
```

We know we need to use a subquery to get 'average_salary', as it involves an aggregate function, the `AVG(salary)`. So let's write the subquery:  

```{sql, eval=FALSE}
SELECT AVG(salary)
FROM employees"
```

Now wherever we see 'average_salary' in the first query, we replace it with the subquery in parentheses

```{sql, connection = db, output.var="out"}
SELECT *
FROM employees
WHERE country = 'Japan' AND salary > (SELECT AVG(salary) FROM employees)
```
```{r, echo=FALSE, eval=TRUE}
out
```

<br>
<blockquote class="task">
**Task - 5 mins**

Write a query using a subquery to answer the following:

<br>
<center>
**"Find all the employees in Legal who earn less than the mean `salary` in that same department."**
</center>
<br>

<details>
<summary>**Solution**</summary>
```{sql, connection = db, output.var="out"}
SELECT *
FROM employees
WHERE department = 'Legal' AND salary < (
  SELECT AVG(salary)
  FROM employees
  WHERE department = 'Legal'
)
```
```{r, echo=FALSE, eval=TRUE}
out
```
</details>
</blockquote>
<br>

<br>
<blockquote class="task">
**Harder Task - 10 mins** 

We're going to use what we've learned in the `GROUP BY` and subqueries sections to answer the following:

<br>
<center>
**"Find all the employees in the United States who work the most common full-time equivalent hours across the corporation."**
</center>
<br>

Think of this as being two separate tasks combined:

* Write a query to find the most common full-time equivalent hours across the corporation
* Use this in a subequery to find those employees in the United States who work those hours

<details>
<summary>**Hints**</summary>
The first query corresponds to the **mode** in statistics: the most commonly occurring value in a set of values. Think of grouping the employees by `fte_hours`, counting the number of employees in each group, and then finding the `fte_hours` with the largest count.
</details>

<details>
<summary>**Solution**</summary>

*Query for the mode*

```{sql, connection = db, output.var="out"}
SELECT fte_hours
FROM employees
GROUP BY fte_hours
HAVING COUNT(*) = (
  SELECT MAX(count) AS max_count
  FROM (
    SELECT COUNT(*) AS count
    FROM employees
    GROUP BY fte_hours
  ) AS temp
)
```
```{r, echo=FALSE, eval=TRUE}
out
```

*Use the mode in subquery*

```{sql, connection = db, output.var="out"}
SELECT *
FROM employees
WHERE country = 'United States' AND fte_hours = (
  SELECT fte_hours
  FROM employees
  GROUP BY fte_hours
  HAVING COUNT(*) = (
    SELECT MAX(count) AS max_count
    FROM (
      SELECT COUNT(*) AS count
      FROM employees
      GROUP BY fte_hours
    ) AS temp
  )
)
```
```{r, echo=FALSE, eval=TRUE}
out
```

```{r}

```


</details>
</blockquote>
<br>

# Recaps

## Order of definition

Now we've completed our discussion of query syntax, here are **all** the different components of a `SELECT` query, the order in which they must appear, and whether they are required or optional <br>

| Order of definition | Keyword | Specifies | Required? |
| --- | --- |--- | --- |
| 1 | `SELECT` | Column to query | Yes |
| 2 | `AS` | Column alias | No |
| 3 | `FROM` | Table to query |  Yes |
| 4 | `WHERE` | Row-level filter | No |
| 5 | `GROUP BY` | Grouping for aggregates | No |
| 6 | `HAVING` | Group-level filter | No |
| 7 | `ORDER BY` | Sort order | No |
| 8 | `LIMIT` | How many records to return | No |

## Useful mnemonic for order of definition

| Keyword | Mnemonic |
| --- | --- |
| `SELECT AS` | So |
| `FROM` | Few |
| `WHERE` | Workers
| `GROUP BY` | Go
| `HAVING` | Home
| `ORDER BY` | On Time

## Order of execution

The order above applies for **writing** `SQL` queries, but it is helpful to understand the **order of execution** of the parts of a `SQL` query. This can help explain why, for example, an alias that you declare in the `SELECT` part can't be used in `HAVING`, and so on...

| Order of execution | Keyword |
| --- | --- |
| 1 | `FROM` |
| 2 | `WHERE` |
| 3 | `GROUP BY` |
| 4 | `HAVING` |
| 5 | `SELECT` |
| 6 | `ORDER BY` |
| 7 | `LIMIT` |



