---
title: "Hypothesis tests & probability - homework"
output: html_notebook
---

```{r}
library(tidyverse)
library(infer)
library(skimr)
data(msleep)
```

1.1 Hypothesis testing - practical

```{r}
sleep_data <- select(msleep, name, sleep_total)
```

```{r}
skim(sleep_data)
sleep_data %>% 
ggplot(aes(x = sleep_total)) +
  geom_histogram(binwidth = 2, color = "white") 

  boxplot(sleep_data$sleep_total)
```
Roughly normal distribution, over 30 samples (83) and independent observations so 1 sample t test is applicable

H0: μsleep_total = 7.0
Hα: μsleep_total ≠ 7.0
α = 0.05
```{r}
t_stat_infer <- sleep_data %>%
  t_stat(response = sleep_total, mu = 7) 

t_stat_infer
```
```{r}
null_distribution_sleep_total <- sleep_data %>% 
  specify(response = sleep_total) %>% 
  hypothesize(null = "point", mu = 7) %>% 
  calculate(stat = "t") 

head(null_distribution_sleep_total)
```

```{r}
null_distribution_sleep_total %>%
  visualize(method = "theoretical") + 
  shade_p_value(obs_stat = t_stat_infer, direction = "both") 
```

```{r}
t_test_results <- sleep_data %>%
  t_test(response = sleep_total, alternative = "two.sided", mu = 7.00)

t_test_results
```
 p < 0.05 so null hypothesis can be rejected and conclude that jabberwockies sleep significantly different from the overall population mean.

#Perform an appropriate statistical test to determine whether omnivores sleep for significantly longer than herbivores, on average.

H0 : μsleep_total(omni)−μsleep_total(herbi)=0
Ha : μsleep_total(omni)−μsleep_total(herbi)<0
α = 0.05  

```{r}
msleep %>%
  group_by(vore) %>%
  summarise(n = n())
```

```{r}
vore_sleep <- msleep %>%
  filter(vore == "omni" | vore == "herbi") 

vore_sleep
```

```{r}
vore_sleep %>%
  ggplot(aes(y = sleep_total, x = vore)) +
  geom_boxplot()
```
Caqlculate null distribution
```{r}
null_distribution_vore <- vore_sleep %>% 
  specify(sleep_total ~ vore) %>% 
  hypothesise(null = "independence") %>% 
  generate(reps = 1000, type = "permute") %>% 
  calculate(stat = "diff in means", order = c("omni", "herbi")) 

head(null_distribution_vore)
```
Observed statistic

```{r}
observed_stat_vore <- vore_sleep %>% 
  specify(sleep_total ~ vore) %>%
  calculate(stat = "diff in means", order = c("omni", "herbi"))
  
observed_stat_vore
```

```{r}
null_distribution_vore %>%
  visualise() +
  shade_p_value(obs_stat = observed_stat_vore, direction = "left")
```
Calculate p value
```{r}
p_value_vore <- null_distribution_vore %>%
  get_p_value(obs_stat = observed_stat_vore, direction = "left")
p_value_vore
```

p > 0.05 so null hypothesis cannot be rejected and conclude omnivores do not sleep significantly longer than herbivores on average

# Perform an appropriate statistical test to determine whether the proportion of domesticated animal types in the population of animal types is greater than 5%.


Assuming animal types refers to omniverous and herbivorous categories from previous question

Create is_domesticated variable
```{r}
domesticated <- vore_sleep %>% 
  mutate(is_domesticated = if_else(conservation == "domesticated" & !is.na(conservation), "yes", "no")) 
```

H0: πall_animal−πdomesticated_animal=0
Ha: πall_animal−πdomesticated_animal≠0
α = 0.05

```{r}
ggplot(domesticated, aes(x = vore, fill = is_domesticated)) +
  geom_bar(position = "fill") +
  coord_flip()
```

```{r}
observed_stat_dom <- domesticated %>% 
  specify(is_domesticated ~ vore, success = "yes") %>% 
  calculate(stat = "diff in props", order = c("herbi", "omni"))

observed_stat_dom
```

```{r}
null_distribution_dom <- domesticated %>%
  specify(is_domesticated ~ vore, success = "yes") %>% 
  hypothesize(null = "independence") %>% 
  generate(reps = 1000, type = "permute") %>%
  calculate(stat = "diff in props", order = c("herbi", "omni"))

head(null_distribution_dom)
```

```{r}
null_distribution_dom %>%
  visualise() +
  shade_p_value(obs_stat = observed_stat_dom, direction = "right")
```

```{r}
p_value_dom <- null_distribution_dom %>%
  get_p_value(obs_stat = observed_stat_dom, direction = "right")

p_value_dom
```

Just realised I've probably just worked out the wrong thing the wrong way,hmmm


#1.2 Hypothesis testing - interpretation
#1.2.1 Defining the Hypothesis

1
H0: πknow of coffee shop >=0.40
Ha: πknow of coffee shop≠0.40

method = simulate

2
H0: πbanner_top−πbanner_right=0
Ha: πbanner_top−πbanner_right>0

method = permutation

3
H0: μpart_width = 145mm
Ha: μpart_width ≠ 145mm

method = bootstrapping

#1.2.2 Interpreting the results

1. Coffee shop problem. Significance level: 0.05, calculated p-value: 0.07
p greater than 0.05 therefore null hypothesis confirmed i.e. not able to confirm less than 40% of population know of the coffee shop.

2. Website company problem. Significance level: 0.01, p-value: 0.006
p less than 0.001 therefore null hypothesis rejected i.e. Users are more likely to click banner at top of the page.

3. Manufacturing company problem. Significance level: 0.05, p-value: 0.55
p more than 0.05 therefore null hypothesis confirmed i.e. manufacturing process has not significantly drifted from 145mm